A Formal Logic for Formal Category Theory

### 2022-10-14 paper/paper.tex
finish proofreading pass over main body

Finally we have several forms of context which are used in the theory.
The contexts $\Gamma$ of term variables with their types are as usual;
we write ${\Gamma \isadtctx}$ to indicate that a context is well-formed.
We name the remaining contexts after the judgements that they are used
by.  The set contexts $\Gamma \vdash \Xi \boundary$, which will be used
to type-check sets, contain object variables with their categories. The
two forms of set context are $\alpha : \cat C$, containing one variable
that can be used both contravariantly and covariantly, and $\alpha :
\cat C ; \beta : \cat D$, containing a contravariant variable $\alpha$
and covariant variable $\beta$.  Finally, the transformation contexts
$\Gamma \vdash \Phi \isavectx$ contain element variables with their sets, alternating with those
sets' object variables with their categories.  A typical $\Phi$ has the
shape
\[
\alpha_1 : \cat C_1, x_1:R_1(\alpha_1,\alpha_2), \alpha_2 : \cat C_2,
x_2:R_2(\alpha_2,\alpha_3), \ldots, R_n(\alpha_n,\alpha_{n+1}),
\alpha_{n+1} : \cat C_{n+1}
\]
and represents the composition of the ``relations'' $R_1, R_2, R_3, \ldots,
R_n$.  We write $d^-(\Phi)$ for the first category variable in $\Phi$
(which we regard as the negative/contravariant position), $d^+(\Phi)$
for the last category variable in $\Phi$ (which we regard as the
positive/covariant position), and $\Phi_1 \jnctx \Phi_2$ for the append
of two category contexts where the last variable in $\Phi_1$ is equal to
the first variable in $\Phi_2$.  Formal inductive definitions are in the
appendix, but intuitively we have: 
\[
\begin{array}{ccl}
d^-(\alpha_1 : \cat C_1, x_1:R_1(\alpha_1,\alpha_2), \ldots, x_n:R_n(\alpha_n,\alpha_n), \alpha_{n+1} : \cat C_{n+1}) & = & \alpha_1 : \cat C_1\\
d^+(\alpha_1 : \cat C_1, x_1:R_1(\alpha_1,\alpha_2), \ldots, x_n:R_n(\alpha_n,\alpha_n), \alpha_{n+1} : \cat C_{n+1}) & = & \alpha_{n+1}:\cat C_{n+1}\\
(\Phi_1,\beta:\cat D) \jnctx (\beta:\cat D, \Phi_2) & = & \Phi_1, \beta : \cat D, \Phi_2
\end{array}
\]

@@

\begin{figure}[t]
  \begin{mathpar}
    \begin{array}{rccl}
      \textrm{Categories} &\cat C, \cat D, \cat E & ::= & \unquoth M \pipe \cat C \times \cat D \pipe \cat 1
      \pipe \graphProf \alpha \beta P \pipe \negPresheaf {\cat C} \pipe \posPresheaf {\cat C}
      \\
      \textrm{Objects} & A, B, C, a, b, c & ::= & \alpha \pipe M a \pipe (a,b)
      \pipe () \pipe \pi_i a \pipe (a_-, a_+, s) \pipe \pi_- a \pipe
      \pi_+ a \pipe \lambda \alpha:\cat C.R \\
      \textrm{Sets} & P,Q,R & ::= & \punitinXfromYtoZ {\cat C} a b \pipe \tensorexistsXwithYandZ {\beta} P Q \pipe \homrallXYtoZ \beta P Q \pipe \homlallXYtoZ \alpha R S 
      \pipe 1 \pipe P \times Q\\
      &&& \pipe \harrapp M a b \pipe \negPresheafAppPtoX a b \pipe \posPresheafAppPtoX a b\\
      \textrm{Elements} & s,t,u & ::= & 
    x \pipe 
    {\punitelimkontZatABC {\alpha. t}{b_1} s {b_2}} \pipe
    \punitrefl b \pipe 
    {\tensorelimWkontZ s {x.\beta.y.r}} \pipe 
    {\tensorintroatXwithYandZ b s t} \pipe
    {\homrappXtoYatZ s t a} \\
    &&&
    \pipe {\homrlambdaXatYdotZ {x}{\alpha} s}\pipe 
    {\homlappXtoYatZ t s a} \pipe 
    {\homllambdaXatYdotZ {x}{\alpha} s} \pipe
    \pi_i s \pipe 
    (s_1,s_2) \pipe
    () \pipe \pi_e a \pipe \pendappXtoY M b \\ 
      \textrm{Type} & A,B,C & ::= &
%      \sum_{X:A} B \pipe \prod_{X:A} B \pipe \Id A M N \\
      \ldots \pipe \smallCats \pipe \Cat \pipe \varr{\cat C}{\cat D}\pipe \harr {\cat C} {\cat D}\pipe \pendallXdotY {\alpha:{\cat C}} R\\
      \textrm{Term} & L,M,N & ::= & 
      %% & X \pipe \lambda X:M. N \pipe M\,N \pipe (M,N) \pipe \pi_1 M \pipe \pi_2 M \mid \text{refl}_{M} \\
      \ldots \pipe \lceil \cat C \rceil \pipe \lambda {\alpha:\cat C}. a \pipe \lambda ({\alpha:\cat C};{\beta: \cat D}). R \mid \lambda \alpha.t \\
      \textrm{Type Context} & \Gamma, \Delta & ::= & \cdot \pipe \Gamma, X : A\\
      \textrm{Set Context} & \Xi, Z & ::= & \alpha:\cat C \pipe \alpha:\cat C;\beta: \cat D\\
      \textrm{Trans. Context} & \Phi,\Psi & ::= & \alpha : \cat C \pipe \Phi , x : P,\beta:\cat D\\
    \end{array}
  \end{mathpar}
  \caption{\vett{} Syntactic Forms}
  \label{fig:syntax}
\end{figure}

@@

The variable rules for objects and elements are 
\[
\inferrule*{}{ \Gamma \pipe \alpha : \cat C \vdash \alpha : \cat C}
\qquad
\inferrule*{}{ \Gamma \pipe \alpha : \cat C, x : R, \beta : \cat D \vdash x : R}
\]
All syntactic forms typed in context admit an action of substitution.
For types and terms, this is as usual. Objects $\alpha:\cat C \vdash a :
\cat D$ can be substituted for object variables $\beta : \cat D$ in
other objects. We can also substitute objects into \emph{sets}, that is,
if we have a set $P$ parameterized by a contravariant variable $\alpha :
\cat C$ and a covariant variable $\beta: \cat D$, then we can substitute
objects $a : \cat C$ and $b : \cat D$ for these variables
$P[a/\alpha;b/\beta]$. This generalizes the ordinary precomposition of a
relation by a function.  Semantically this is the ``restriction'' of a
profunctor along two functors, which is just composition of functors if
a profunctor is viewed as a functor to $\Set$. Modeling this operation
as a substitution considerably simplifies reasoning using profunctors.
%
Finally we have the action of substitution on elements/natural transformations.
First, we can substitute elements/natural transformations for the set
variables in elements, denoting the composition of natural
transformations.  Second, an element is also parameterized
by a contravariant and a covariant category variable $\alpha;\beta$. We
can think of natural transformations as \emph{polymorphic} in the
categories involved, and so when we make a transformation substitution,
we also \emph{instantiate} the polymorphic category variables with
objects.
%
The full syntactic details of substitution are included in the appendix.

@@

In this section we discuss some connectives for constructing categories,
which are specified by introduction and elimination rules in
Figure~\ref{fig:category-connectives} (the $\beta\eta$ equality and
substitution rules are included in the appendix).  The introduction and
elimination rules make use of functors, profunctors, and natural
transformations.
%
First we introduce the additives: the unit category $1$ and product
category $\cat C \times \cat D$ have the usual introduction and
elimination rules defining functors to/from them.  
%
Next, we introduce the \emph{graph of a profunctor} $\graphProf \alpha
\beta P$. Just as a relation $R : A \times B \to \Set$ can be viewed
as a subset $\{ (a,b) \in A\times B | R(a,b)\}$, any profunctor $P :
C_-^o\times C_+ \to \Set$ can be viewed as a category with a functor
to $C_- \times C_+$ (no op), specifically a two-sided discrete
fibration. In set-based category theory, the objects are triples
$(a_-, a_+, s : P(a_-,a_+))$ and morphisms from $(a_-,a_+, s)$ to
$(a_-',a_+', s')$ are pairs of morphisms $f_- : a_- \to a_-'$ and $f_+
: a_+ \to a_+'$ such that $P(\id,f_+)(s) = P(f_-,\id)(s')$.
%
With various choices of $P$, this connective can be used to define the
arrow category, slice category and comma category constructions.
%
In our syntax we define it as the universal category $\cat C$ equipped
with functors to $\cat C_-$ and $\cat C_+$ and a morphism to $P$.

@@

Next, in Figure~\ref{fig:set-connectives}, we cover the connectives for
the sets/profunctors, which classify elements/natural transformations
(the $\beta/eta$-rules are in the appendix).
%% %
%% The top of the figure includes the formation rules for tensor products
%% $\tensorexistsXwithYandZ \beta P Q$, the covariant $\homrallXYtoZ
%% \alpha R P$ and contravariant homs $\homlallXYtoZ {\alpha} R P$ and
%% the unit type $\punitinXfromYtoZ {\cat C} {a_1}{a_2}$, and
%% the cartesian unit and product types $1$ and $R \times S$.  
%% %
%% The remainder of the figure gives the introduction and elimination rules
%% for these sets .  
%% %
First, the unit set is our syntax for the profunctor of morphisms of a
category $\cat C$, which is well-formed when given two objects of the
same category.  
%
Its introduction and elimination rules are analogous to the usual rules
for equality in Martin-L\"of type theory.  The introduction rule is the
identity morphism (reflexivity) and the elimination rule is an induction
principle: we can use a term of $s : \punitinXfromYtoZ {\cat C} A B$ by
specifying the behavior when $s$ is of the form $\punitrefl \alpha$ in
the form of a continuation $\alpha. t$.  Like the $J$ elimination rule
for equality in Martin-L\"of type theory, $P$ must be ``fully general'',
i.e. well-typed for variables $\alpha$ and $\beta$; this is because for
distinct variables $\alpha$ and $\beta$, $\punitinXfromYtoZ{\cat
  C}{\alpha}{\beta}$ denotes the unit, which has a universal property,
but $\punitinXfromYtoZ{\cat C} A B$ denotes a restriction of a unit,
which in general does not.  Those familiar with linear
logic~\cite{polakow-pfenning} might expect a more general rule, with the
continuation $t$ typed in a context $\Phi_l \jnctx \Phi_r$ and $\Phi_l
\jnctx \Phi \jnctx \Phi_r$ as the context of the conclusion of the rule.
Because of dependency, this is not necessarily well-formed in cases
where the endpoints $A$ and $B$ of $\punitinXfromYtoZ{\cat C} A B$ are
not distinct variables.  However, the instances of this more general
rule that do type check are derivable from our more restricted rule
using right/left-hom types.

@@

    \text{Unit/morphism set:}\\
    \inferrule
  {\Gamma \pipe d^-\Xi \vdash a_1 : \cat C\\\\ \Gamma \pipe d^+ \Xi \vdash a_2 : \cat C}
  {\Gamma \pipe \Xi \vdash \punitinXfromYtoZ {\cat C} {a_1} {a_2} \isaSet}
  \quad
    \inferrule
    {\Gamma\pipe\beta: \cat D \vdash a : \cat C}
    {\Gamma \pipe \beta : \cat D \vdash \punitrefl a : \punitinXfromYtoZ {\cat C} a a}
    \quad
%%     \inferrule*[right=Unit Elim]
%%     {\Gamma\vdash Q : \harr {\cat C} {\cat C}\and
%%      \Gamma\vdash t : \pendallXdotY {\alpha:\cat C} {\harrapp Q \alpha \alpha}}
%%     {\Gamma \vdash \punitelimtoYwithkontZ Q N :
%%       \pendallXdotY {\alpha_1} {\homrallXYtoZ {\alpha_2} {\punitinXfromYtoZ {\cat C} {\alpha_1}{\alpha_2}} {\harrapp Q {\alpha_1}{\alpha_2}}}}
    \inferrule
    {\Gamma\pipe \alpha:\cat C; \beta:\cat C \vdash P \isaSet \\\\
     \Gamma\pipe \alpha:\cat C\vdash t : P[\alpha/\alpha;\alpha/\beta]\\\\
     \Gamma\pipe\Phi \vdash s : \punitinXfromYtoZ {\cat C} A B}
    {\Gamma \pipe \Phi \vdash \punitelimkontZatABC {\alpha. t} A s B : P[A/\alpha;B/\beta]}
    \\
  %% \inferrule*[right=Tensor intro function]
  %% {}
  %% {\alpha : \cat C, q : Q, \beta : \cat D, r : R, \gamma : \cat E \vdash (q, \beta,r) : \exists \beta : \cat D. Q \odot R}

@@

Next, we can prove that two definitions of an adjunction are equivalent:
\begin{lemma} \label{lem:adjunction}
  For $R : \varr {\cat D}{\cat C}$ and $L : \varr {\cat
    C}{\cat D}$, the following are in bijection:
  \begin{enumerate}
  \item An isomorphism of profunctors $(\punitinXfromYtoZ {\cat D} {L \alpha}{\beta}) \cong (\punitinXfromYtoZ {\cat C} \alpha {R \beta})$
  \item A unit $\eta : \pendallXdotY \alpha {\punitinXfromYtoZ {\cat C} \alpha
    {R(L\alpha)}}$ and co-unit $\varepsilon: \pendallXdotY \beta
    {\punitinXfromYtoZ {\cat D} {L(R(\beta))}{\beta}}$ satisfying triangle
    identities.
  \end{enumerate}
\end{lemma}
\begin{proof}
  Given the forward homomorphism $\text{lr}$, we can construct $\eta =
  \pendlambdaXdotY \alpha \appunary {\text{lr}} \alpha {\punitrefl
    \alpha} {L \alpha}$ and given the unit we can reconstruct the
  forward homomorphism using
  $\text{comp}$ and $\text{fctor}$ as defined in
  Construction\ref{construction:synthetic-composition} as
  $\appbinary {\text{comp}} \alpha
  {\pendappXtoY \eta \alpha} {R(L\alpha)} {(\appunary
    {\text{fctor}(R)}{L \alpha} {f} {\beta})} {R \beta}$.
\end{proof}

\noindent We can define weighted limits, which include ordinary limits and
Kan extensions.
\begin{definition}
  For a functor $D : \varr {\cat J}{\cat C}$ and a profunctor $W :
  \harr{\cat K}{\cat J}$, the limit of $D$ weighted by $W$ is (if it
  exists) a functor $\weightedLimitDW D W : \varr{\cat K}{\cat C}$
  with an isomorphism $\punitinXfromYtoZ {\cat C} {\alpha} {(\weightedLimitDW D W) k} \cong \homrallXYtoZ j {W k j} {(\punitinXfromYtoZ {\cat C} \alpha {D j})}$
\end{definition}
This generalizes the usual definition that a morphism into a limit is
a cone over the diagram $(\punitinXfromYtoZ {\cat C} \alpha {D j})$ to
be parameterized by a weight $W k j$.
Then we can prove the well-known theorem that right adjoints preserve (weighted) limits:
\begin{theorem}
  If $\weightedLimitDW D W$ exists and is a limit and $R : \varr{\cat
    C}{\cat C'}$ has a left adjoint $L$, then $\lambda
  \kappa. R((\weightedLimitDW D W) \kappa)$ is the limit of $\lambda
  j. R(D j)$ weighted by $W$.
\end{theorem}
\begin{proof}
\[
    \punitinXfromYtoZ {} {\gamma} {R ((\weightedLimitDW D W) \kappa)}
    \cong \punitinXfromYtoZ {}{L \gamma} {(\weightedLimitDW D W) \kappa}
    \cong \homrallXYtoZ {j}{W k j} {\punitinXfromYtoZ {}{L \gamma}{D j}}
    \cong \homrallXYtoZ {j}{W k j} {\punitinXfromYtoZ {}{\gamma}{R(D j)}}\\
\]
This is a high level proof in terms of isomorphisms that may be written
in VETT. The first two steps are the instantiation of assumptions
(adjointness, weighted limits). The last step uses the fact that a
natural isomorphisms lift to natural isomorphism of homs of profunctors.
The construction of this isomorphism illustrates how naturality need not
be proved explicitly in \vett{}.  For any $\phi: \homunary \alpha {R'
  \alpha \beta} \beta {R \alpha \beta}$ and $\psi : \homunary \gamma {S
  \gamma \beta} \beta {S' \gamma \beta}$ we can construct a natural
transformation $\phi \triangleright \psi : \hombinary \gamma
{(\homrallXYtoZ \beta {R \alpha \beta}{S \gamma \beta})} {\alpha}
{R'\alpha\beta} \beta {S'\gamma\beta}$ as \\ $\lambdabinary {\gamma} {f}
{\alpha} r \beta {\appunary \psi \gamma {(\homrappXtoYatZ f {(\appunary
      \phi \alpha r \beta)} \beta)} \beta}$. Furthermore if $\phi$ and
$\psi$ have inverses, then $\phi^{-1} \triangleright \psi^{-1}$ is the
inverse of $\phi \triangleright \psi$.
\end{proof}

@@

In the presence of restrictions, every 2-cell can be represented as a
``globular'' 2-cell where the left and right frame are identities. For
example the 2-cell $\phi$ above can be represented as one with the
same domain but whose codomain is $S(f,g)$. This property is crucial
for the completeness of our semantics as we only include a syntax for these globular terms 
%
Each component of this definition has a direct correspondence to a
syntactic structure in \ohol{}. The objects of $\mathcal V_o$ models
the category judgment and the morphisms model the functor
judgment. The set $\mathcal V_h$ models the profunctor judgment. A
composable string $R_0 \cdots R_n$ models the profunctor contexts. The
2-cells correspond to the natural transformation judgment where we
have taken the restriction $S(F,G)$ of the codomain.

>>>>

Finally we have several forms of context which are used in the theory.
The contexts $\Gamma$ of term variables with their types are as usual;
we write ${\Gamma \isadtctx}$ to indicate that a context is well-formed.
We name the remaining contexts after the judgements that they are used
by.  The set contexts $\Gamma \vdash \Xi \boundary$, which will be used
to type-check sets, contain object variables with their categories. The
two forms of set context are $\alpha : \cat C$, containing one variable
that can be used both contravariantly and covariantly, and $\alpha :
\cat C ; \beta : \cat D$, containing a contravariant variable $\alpha$
and covariant variable $\beta$.  Finally, the transformation contexts
$\Gamma \vdash \Phi \isavectx$ contain element variables with their sets, alternating with those
sets' object variables with their categories.  A typical $\Phi$ has the
shape
\[
\alpha_1 : \cat C_1, x_1:R_1(\alpha_1,\alpha_2), \alpha_2 : \cat C_2,
x_2:R_2(\alpha_2,\alpha_3), \ldots, R_n(\alpha_n,\alpha_{n+1}),
\alpha_{n+1} : \cat C_{n+1}
\]
and represents the composition of the ``relations'' $R_1, R_2, R_3, \ldots,
R_n$.  We write $d^-(\Phi)$ for the first category variable in $\Phi$
(which we regard as the negative or contravariant position), $d^+(\Phi)$
for the last category variable in $\Phi$ (which we regard as the
positive or covariant position), and $\Phi_1 \jnctx \Phi_2$ for the append
of two transformation contexts where the last variable in $\Phi_1$ is equal to
the first variable in $\Phi_2$.  Formal inductive definitions are in the
appendix, but intuitively: 
\[
\begin{array}{ccl}
d^-(\alpha_1 : \cat C_1, x_1:R_1(\alpha_1,\alpha_2), \ldots, x_n:R_n(\alpha_n,\alpha_n), \alpha_{n+1} : \cat C_{n+1}) & = & \alpha_1 : \cat C_1\\
d^+(\alpha_1 : \cat C_1, x_1:R_1(\alpha_1,\alpha_2), \ldots, x_n:R_n(\alpha_n,\alpha_n), \alpha_{n+1} : \cat C_{n+1}) & = & \alpha_{n+1}:\cat C_{n+1}\\
(\Phi_1,\beta:\cat D) \jnctx (\beta:\cat D, \Phi_2) & = & \Phi_1, \beta : \cat D, \Phi_2
\end{array}
\]

@@

\begin{figure}[t]
  \begin{mathpar}
    \begin{array}{rccl}
      \textrm{Categories} &\cat C, \cat D, \cat E & ::= & \unquoth M \pipe \cat C \times \cat D \pipe \cat 1
      \pipe \graphProf \alpha \beta P \pipe \negPresheaf {\cat C} \pipe \posPresheaf {\cat C}
      \\
      \textrm{Objects} & a, b, c & ::= & \alpha \pipe M a \pipe (a,b)
      \pipe () \pipe \pi_i a \pipe (a_-, a_+, s) \pipe \pi_- a \pipe
      \pi_+ a \pipe \lambda \alpha:\cat C.R \\
      \textrm{Sets} & P,Q,R & ::= & \punitinXfromYtoZ {\cat C} a b \pipe \tensorexistsXwithYandZ {\beta} P Q \pipe \homrallXYtoZ \beta P Q \pipe \homlallXYtoZ \alpha R S 
      \pipe 1 \pipe P \times Q\\
      &&& \pipe \harrapp M a b \pipe \negPresheafAppPtoX a b \pipe \posPresheafAppPtoX a b\\
      \textrm{Elements} & s,t,u & ::= & 
    x \pipe 
    {\punitelimkontZatABC {\alpha. t}{b_1} s {b_2}} \pipe
    \punitrefl b \pipe 
    {\tensorelimWkontZ s {x.\beta.y.r}} \pipe 
    {\tensorintroatXwithYandZ b s t} \pipe
    {\homrappXtoYatZ s t a} \\
    &&&
    \pipe {\homrlambdaXatYdotZ {x}{\alpha} s}\pipe 
    {\homlappXtoYatZ t s a} \pipe 
    {\homllambdaXatYdotZ {x}{\alpha} s} \pipe
    \pi_i s \pipe 
    (s_1,s_2) \pipe
    () \pipe \pi_e a \pipe \pendappXtoY M b \\ 
      \textrm{Type} & A,B,C & ::= &
%      \sum_{X:A} B \pipe \prod_{X:A} B \pipe \Id A M N \\
      \ldots \pipe \smallCats \pipe \Cat \pipe \varr{\cat C}{\cat D}\pipe \harr {\cat C} {\cat D}\pipe \pendallXdotY {\alpha:{\cat C}} R\\
      \textrm{Term} & L,M,N & ::= & 
      %% & X \pipe \lambda X:M. N \pipe M\,N \pipe (M,N) \pipe \pi_1 M \pipe \pi_2 M \mid \text{refl}_{M} \\
      \ldots \pipe \lceil \cat C \rceil \pipe \lambda {\alpha:\cat C}. a \pipe \lambda ({\alpha:\cat C};{\beta: \cat D}). R \mid \lambda \alpha.t \\
      \textrm{Type Context} & \Gamma, \Delta & ::= & \cdot \pipe \Gamma, X : A\\
      \textrm{Set Context} & \Xi, Z & ::= & \alpha:\cat C \pipe \alpha:\cat C;\beta: \cat D\\
      \textrm{Trans. Context} & \Phi,\Psi & ::= & \alpha : \cat C \pipe \Phi , x : P,\beta:\cat D\\
    \end{array}
  \end{mathpar}
  \caption{\vett{} Syntactic Forms}
  \label{fig:syntax}
\end{figure}

@@

The variable rules for objects and elements are 
\[
\inferrule*{}{ \Gamma \pipe \alpha : \cat C \vdash \alpha : \cat C}
\qquad
\inferrule*{}{ \Gamma \pipe \alpha : \cat C, x : R, \beta : \cat D \vdash x : R}
\]
As in linear logic, the latter rule applies only when the context
contains a single set $R$.  All syntactic forms typed in context admit
an action of substitution.  For types and terms, this is as
usual. Objects $\alpha:\cat C \vdash a : \cat D$ can be substituted for
object variables $\beta : \cat D$ in other objects. We can also
substitute objects into \emph{sets}, that is, if we have a set $P$
parameterized by a contravariant variable $\alpha : \cat C$ and a
covariant variable $\beta: \cat D$, then we can substitute objects $a :
\cat C$ and $b : \cat D$ for these variables $P[a/\alpha;b/\beta]$. This
generalizes the ordinary precomposition of a relation by a function.
Semantically this is the ``restriction'' of a profunctor along two
functors, which is just composition of functors if a profunctor is
viewed as a functor to $\Set$. Modeling this operation as a substitution
considerably simplifies reasoning using profunctors.
%
Finally we have the action of substitution on elements/natural transformations.
First, we can substitute elements/natural transformations for the set
variables in elements, denoting the composition of natural
transformations.  Second, an element is also parameterized
by a contravariant and a covariant category variable $\alpha;\beta$. We
can think of natural transformations as \emph{polymorphic} in the
categories involved, and so when we make a transformation substitution,
we also \emph{instantiate} the polymorphic category variables with
objects.
%
The full syntactic details of substitution are included in the appendix.

@@

In this section we discuss some connectives for constructing categories,
which are specified by introduction and elimination rules in
Figure~\ref{fig:category-connectives} (the $\beta\eta$ equality and
substitution rules are included in the appendix).  The introduction and
elimination rules make use of functors, profunctors, and natural
transformations.
%
First we introduce the additives: the unit category $1$ and product
category $\cat C \times \cat D$ have the usual introduction and
elimination rules defining functors to/from them.  
%
Next, we introduce the \emph{graph of a profunctor} $\graphProf \alpha
\beta P$. Just as a relation $R : A \times B \to \Set$ can be viewed
as a subset $\{ (a,b) \in A\times B | R(a,b)\}$, any profunctor $P :
C_-^o\times C_+ \to \Set$ can be viewed as a category with a functor
to $C_- \times C_+$ (no op), specifically a two-sided discrete
fibration. In set-based category theory, the objects are triples
$(a_-, a_+, s : P(a_-,a_+))$ and morphisms from $(a_-,a_+, s)$ to
$(a_-',a_+', s')$ are pairs of morphisms $f_- : a_- \to a_-'$ and $f_+
: a_+ \to a_+'$ such that $P(\id,f_+)(s) = P(f_-,\id)(s')$.
%
With various choices of $P$, this connective can be used to define the
arrow category, slice category and comma category constructions.
%
In our syntax we define it as the universal category $\cat C$ equipped
with functors to $\cat C_-$ and $\cat C_+$ and a natural transformation to $P$.

@@

Next, in Figure~\ref{fig:set-connectives}, we cover the connectives for
the sets/profunctors, which classify elements/natural transformations
(the $\beta/\eta$-rules are in the appendix).
%% %
%% The top of the figure includes the formation rules for tensor products
%% $\tensorexistsXwithYandZ \beta P Q$, the covariant $\homrallXYtoZ
%% \alpha R P$ and contravariant homs $\homlallXYtoZ {\alpha} R P$ and
%% the unit type $\punitinXfromYtoZ {\cat C} {a_1}{a_2}$, and
%% the cartesian unit and product types $1$ and $R \times S$.  
%% %
%% The remainder of the figure gives the introduction and elimination rules
%% for these sets .  
%% %
First, the unit set is our syntax for the profunctor of morphisms of a
category $\cat C$, which is well-formed when given two objects of the
same category.  
%
Its introduction and elimination rules are analogous to the usual rules
for equality in intensional Martin-L\"of type theory.  The introduction
rule is the identity morphism (reflexivity) and the elimination rule is
an induction principle: we can use a term of $s : \punitinXfromYtoZ
{\cat C} a b$ by specifying the behavior when $s$ is of the form
$\punitrefl \alpha$ in the form of a continuation $\alpha. t$.  Like the
$J$ elimination rule for equality in Martin-L\"of type theory, $P$ must
be ``fully general'', i.e. well-typed for variables $\alpha$ and
$\beta$; this is because for distinct variables $\alpha$ and $\beta$,
$\punitinXfromYtoZ{\cat C}{\alpha}{\beta}$ denotes the unit, which has a
universal property, but $\punitinXfromYtoZ{\cat C} a b$ denotes a
restriction of a unit, which in general does not.  Those familiar with
linear logic as in e.g.~\cite{polakow-pfenning} might expect a more
general rule, with the continuation $t$ typed in a context $\Phi_l
\jnctx \Phi_r$, and $\Phi_l \jnctx \Phi \jnctx \Phi_r$ as the context of
the conclusion of the rule.  Because of dependency, this is not
necessarily well-formed in cases where the endpoints $a$ and $b$ of
$\punitinXfromYtoZ{\cat C} a b$ are not distinct variables.  However,
the instances of this more general rule that do type check are derivable
from our more restricted rule using right/left-hom types.

@@

    \text{Unit/morphism set:}\\
    \inferrule
  {\Gamma \pipe d^-\Xi \vdash a_1 : \cat C\\\\ \Gamma \pipe d^+ \Xi \vdash a_2 : \cat C}
  {\Gamma \pipe \Xi \vdash \punitinXfromYtoZ {\cat C} {a_1} {a_2} \isaSet}
  \quad
    \inferrule
    {\Gamma\pipe\beta: \cat D \vdash a : \cat C}
    {\Gamma \pipe \beta : \cat D \vdash \punitrefl a : \punitinXfromYtoZ {\cat C} a a}
    \quad
%%     \inferrule*[right=Unit Elim]
%%     {\Gamma\vdash Q : \harr {\cat C} {\cat C}\and
%%      \Gamma\vdash t : \pendallXdotY {\alpha:\cat C} {\harrapp Q \alpha \alpha}}
%%     {\Gamma \vdash \punitelimtoYwithkontZ Q N :
%%       \pendallXdotY {\alpha_1} {\homrallXYtoZ {\alpha_2} {\punitinXfromYtoZ {\cat C} {\alpha_1}{\alpha_2}} {\harrapp Q {\alpha_1}{\alpha_2}}}}
    \inferrule
    {\Gamma\pipe \alpha:\cat C; \beta:\cat C \vdash P \isaSet \\\\
     \Gamma\pipe \alpha:\cat C\vdash t : P[\alpha/\alpha;\alpha/\beta]\\\\
     \Gamma\pipe\Phi \vdash s : \punitinXfromYtoZ {\cat C} a b}
    {\Gamma \pipe \Phi \vdash \punitelimkontZatABC {\alpha. t} A s B : P[a/\alpha;b/\beta]}
    \\
  %% \inferrule*[right=Tensor intro function]
  %% {}
  %% {\alpha : \cat C, q : Q, \beta : \cat D, r : R, \gamma : \cat E \vdash (q, \beta,r) : \exists \beta : \cat D. Q \odot R}

@@

Next, we can prove that two definitions of an adjunction are equivalent:
\begin{lemma} \label{lem:adjunction}
  For $R : \varr {\cat D}{\cat C}$ and $L : \varr {\cat
    C}{\cat D}$, the following are in bijection:
  \begin{enumerate}
  \item An isomorphism of profunctors $(\punitinXfromYtoZ {\cat D} {L \alpha}{\beta}) \cong (\punitinXfromYtoZ {\cat C} \alpha {R \beta})$
  \item A unit $\eta : \pendallXdotY \alpha {\punitinXfromYtoZ {\cat C} \alpha
    {R(L\alpha)}}$ and co-unit $\varepsilon: \pendallXdotY \beta
    {\punitinXfromYtoZ {\cat D} {L(R(\beta))}{\beta}}$ satisfying triangle
    identities.
  \end{enumerate}
\end{lemma}
\begin{proof}
  Given the forward homomorphism $\text{lr}$, we can construct $\eta =
  \pendlambdaXdotY \alpha \appunary {\text{lr}} \alpha {\punitrefl
    \alpha} {L \alpha}$.  Given the unit we can reconstruct the
  forward homomorphism using
  $\text{comp}$ (composition) and $\text{fctor}$ (functoriality) from
  Construction~\ref{construction:synthetic-composition} as\\
  $\appbinary {\text{comp}} \alpha
  {\pendappXtoY \eta \alpha} {R(L\alpha)} {(\appunary
    {\text{fctor}(R)}{L \alpha} {f} {\beta})} {R \beta}$.
\end{proof}

We can define weighted limits, which as special cases include ordinary
limits and Kan extensions.
\begin{definition}
  For a functor $D : \varr {\cat J}{\cat C}$ and a profunctor $W :
  \harr{\cat K}{\cat J}$, the limit of $D$ weighted by $W$ is (if it
  exists) a functor $\weightedLimitDW D W : \varr{\cat K}{\cat C}$
  with an isomorphism $\punitinXfromYtoZ {\cat C} {\alpha} {(\weightedLimitDW D W) k} \cong \homrallXYtoZ j {W k j} {(\punitinXfromYtoZ {\cat C} \alpha {D j})}$
\end{definition}
This generalizes the usual definition that a morphism into a limit is
a cone over the diagram $(\punitinXfromYtoZ {\cat C} \alpha {D j})$ to
be parameterized by a weight $W k j$.
Then we can prove the well-known theorem that right adjoints preserve (weighted) limits:
\begin{theorem}
  If $\weightedLimitDW D W$ exists and is a limit and $R : \varr{\cat
    C}{\cat C'}$ has a left adjoint $L$, then $\lambda
  \kappa. R((\weightedLimitDW D W) \kappa)$ is the limit of $\lambda
  j. R(D j)$ weighted by $W$.
\end{theorem}
\begin{proof}
\[
    \punitinXfromYtoZ {} {\gamma} {R ((\weightedLimitDW D W) \kappa)}
    \cong \punitinXfromYtoZ {}{L \gamma} {(\weightedLimitDW D W) \kappa}
    \cong \homrallXYtoZ {j}{W k j} {\punitinXfromYtoZ {}{L \gamma}{D j}}
    \cong \homrallXYtoZ {j}{W k j} {\punitinXfromYtoZ {}{\gamma}{R(D j)}}\\
\]
This is a high level proof in terms of isomorphisms that may be written
in VETT. The first two steps are the instantiation of assumptions
(adjointness, weighted limits). The last step uses the fact that a
natural isomorphisms lift to natural isomorphism of homs of profunctors.
The construction of this isomorphism illustrates how naturality need not
be proved explicitly in \vett{}.  For any $\phi: \homunary \alpha {R'
  \alpha \beta} \beta {R \alpha \beta}$ and $\psi : \homunary \gamma {S
  \gamma \beta} \beta {S' \gamma \beta}$ we can construct a natural
transformation $\phi \triangleright \psi : \hombinary \gamma
{(\homrallXYtoZ \beta {R \alpha \beta}{S \gamma \beta})} {\alpha}
{R'\alpha\beta} \beta {S'\gamma\beta}$ as \\ $\lambdabinary {\gamma} {f}
{\alpha} r \beta {\appunary \psi \gamma {(\homrappXtoYatZ f {(\appunary
      \phi \alpha r \beta)} \beta)} \beta}$. Furthermore if $\phi$ and
$\psi$ have inverses, then $\phi^{-1} \triangleright \psi^{-1}$ is the
inverse of $\phi \triangleright \psi$.
\end{proof}

@@

In the presence of restrictions, every 2-cell can be represented as a
``globular'' 2-cell where the left and right frame are identities. For
example the 2-cell $\phi$ above can be represented as one with the
same domain but whose codomain is $S(f,g)$. This property is crucial
for the completeness of our semantics as we only include a syntax for these globular terms.
%
Each component of this definition has a direct correspondence to a
syntactic structure in \ohol{}. The objects of $\mathcal V_o$ models
the category judgment and the morphisms model the functor
judgment. The set $\mathcal V_h$ models the profunctor judgment. A
composable string $R_0 \cdots R_n$ models the profunctor contexts. The
2-cells correspond to the natural transformation judgment where we
have taken the restriction $S(F,G)$ of the codomain.
